Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-17 draft.tex
Minor Changes + ToDo for categorization

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
}

@@

In this paper, we present a thorough analysis of the image-derived data of a cohort study including lumbar back spine datasets.
%%
We analyze their discriminating power for different characteristics of back pain.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R-Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
For example, the analysis of the influence of nutrition on prostate cancer is strongly age dependent and requires therefore age-normalization.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim at collecting as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

>>>>

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
}

@@

In this paper, we present a thorough analysis of the image-derived data of a cohort study including lumbar back spine datasets.
%%
We analyze their discriminating power for different characteristics of back pain.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
For example, the analysis of the influence of nutrition on prostate cancer is strongly age dependent and requires therefore age-normalization.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim at collecting as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%
